fGFPCA dynamic prediction (In-sample)

### 2023-01-13 f-GFPCA.Rmd
bias of score

---
title: "fGFPCA dynamic prediction (In-sample)"
output: 
  html_document:
    toc: true
    toc_float: 
      collapsed: false
    toc_depth: 3
date: "2022-11-03"
---

@@

library(here)
library(tidyverse)
library(ggplot2)
library(fcr)
library(refund)
library(lme4)
library(ggpubr)
set.seed(1025)
```

@@

@@

in_pred_clean <- function(subjID){
  # make predictions
  pred_t195 <- in_samp_dyn_pred(fpca_fit, df_pred_unique %>% filter(id==subjID & bin<=195))
  pred_t395 <- in_samp_dyn_pred(fpca_fit, df_pred_unique %>% filter(id==subjID & bin<=395))
  pred_t595 <- in_samp_dyn_pred(fpca_fit, df_pred_unique %>% filter(id==subjID & bin<=595))

  df_pred_t195 <- data.frame(mid,
                             eta=pred_t195$eta_pred,
                             eta_std = pred_t195$eta_std_var) %>%
    mutate(eta_lower = eta-eta_std, 
             eta_upper = eta+eta_std) %>% 
    filter(mid > 195)

  df_pred_t395 <- data.frame(mid,
                             eta=pred_t395$eta_pred,
                             eta_std = pred_t395$eta_std_var) %>%
    mutate(eta_lower = eta-eta_std, 
             eta_upper = eta+eta_std) %>% 
    filter(mid > 395)

  df_pred_t595 <- data.frame(mid,
                             eta=pred_t595$eta_pred,
                             eta_std = pred_t595$eta_std_var) %>%
    mutate(eta_lower = eta-eta_std, 
             eta_upper = eta+eta_std) %>% 
  filter(mid > 595)

  return(list(df_pred_t195, df_pred_t395, df_pred_t595))
}

@@

```{r, fig.width=12, fig.height=4}
# Try on subjects 1-4
bind_rows(in_pred_clean(1), .id = "type") %>%
  mutate(type=factor(type, levels=1:3, 
                     labels = c("Up to 195", "Up to 395", "Up to 595"))) %>%
  left_join(df_true_eta %>% filter(id==1), by = "mid") %>% 
  ggplot()+
  geom_line(aes(x=mid, y = true, col = "True"))+
  geom_line(aes(x=mid, y = eta), na.rm = T)+
  geom_line(aes(x=mid, y=eta_lower), linetype = "dashed", na.rm=T)+
  geom_line(aes(x=mid, y = eta_upper), linetype = "dashed", na.rm=T)+
  labs(title = "Subject 1", x = "bin", y="")+
  facet_wrap(~type)

bind_rows(in_pred_clean(2), .id = "type") %>%
  mutate(type=factor(type, levels=1:3, 
                     labels = c("Up to 195", "Up to 395", "Up to 595"))) %>%
  left_join(df_true_eta %>% filter(id==2), by = "mid") %>% 
  ggplot()+
  geom_line(aes(x=mid, y = true, col = "True"))+
  geom_line(aes(x=mid, y = eta), na.rm = T)+
  geom_line(aes(x=mid, y=eta_lower), linetype = "dashed", na.rm=T)+
  geom_line(aes(x=mid, y = eta_upper), linetype = "dashed", na.rm=T)+
  labs(title = "Subject 2", x = "bin", y="")+
  facet_wrap(~type)
```

## Some clarification points

1. When fitting local GLMMs, it seems that the intervals are essentially independent. We do not consider the effect of neighborhood intervals on a specific interval.

@@

#### In- and out-of-sample prediction of latent function

@@

```{r}
# out-of-sample
pred1_t195 <- out_samp_dyn_pred(df_new = df %>% filter(id == 1 & sind_inx <= 195) %>%
                                  select(-eta_i),
                                fpca_fit = fpca_fit)
pred1_t395 <- out_samp_dyn_pred(df_new = df %>% filter(id == 1 & sind_inx <= 395) %>%
                                  select(-eta_i),
                                fpca_fit = fpca_fit)

pred1_t595 <- out_samp_dyn_pred(df_new = df %>% filter(id == 1 & sind_inx <= 595) %>%
                                  select(-eta_i),
                                fpca_fit = fpca_fit)

df_out_pred1 <- data.frame(mid, pred1_t195,pred1_t395, pred1_t595)
df_out_pred1$pred1_t195[1:which(mid==195)] <- NA
df_out_pred1$pred1_t395[1:which(mid==395)] <- NA
df_out_pred1$pred1_t595[1:which(mid==595)] <- NA
df_out_pred1$true <- df[df$id==1 & df$sind_inx%in%mid, "eta_i"]

@@

```{r}
# in-sample  
df_in_pred1 <- data.frame(mid,
                          pred_t195 = in_samp_dyn_pred(fpca_fit, df_pred_unique %>% filter(id==1 & bin<=195))$eta_pred,
                          pred_t395 = in_samp_dyn_pred(fpca_fit, df_pred_unique %>% filter(id==1 & bin<=395))$eta_pred,
                          pred_t595 = in_samp_dyn_pred(fpca_fit, df_pred_unique %>% filter(id==1 & bin<=595))$eta_pred)

df_in_pred1$pred_t195[1:which(mid==195)] <- NA
df_in_pred1$pred_t395[1:which(mid==395)] <- NA
df_in_pred1$pred_t595[1:which(mid==595)] <- NA
df_in_pred1$true <- df[df$id==1 & df$sind_inx%in%mid, "eta_i"]

@@

```{r}
# compare point prediction with in-sample

## out-of-sample
ggplot(df_out_pred1)+
  geom_line(aes(x=mid, y=true, col="True"))+
  geom_line(aes(x=mid, y=pred1_t195, col="Up to 195"), na.rm = T)+
  geom_line(aes(x=mid, y=pred1_t395, col="Up to 395"), na.rm = T)+
  geom_line(aes(x=mid, y=pred1_t595, col="Up to 595"), na.rm = T)+
  labs(y = "Out-of-sample prediction", x = "Binned time")

## in-sample
ggplot(df_in_pred1)+
  geom_line(aes(x=mid, y=true, col="True"))+
  geom_line(aes(x=mid, y=pred_t195, col="Up to 195"), na.rm = T)+
  geom_line(aes(x=mid, y=pred_t395, col="Up to 395"), na.rm = T)+
  geom_line(aes(x=mid, y=pred_t595, col="Up to 595"), na.rm = T)+
  labs(y = "In-sample prediction", x = "Binned time")

@@

- Repeat this for more subjects
- Interval prediction for MLE
- Bias of score for both in-sample and out-of-sample estimation

@@

1. Write out the formula for the BLUPs in the local GLMMs. Since we have a random intercept only model this should not be too complicated. (this is for out of sample prediction only)

2. Approximate these values numerically via numeric integration (like we did witht he AUC paper) and/or explicit laplace approximation 
3. Evaluate performance of dynamic predictions (both using "in sample" data -- those individuals whose BLUPs we can extract directly from the local glmm fits and out of sample data)
4. Write an R package (see Julia's fastGFPCA and Andrew's fcr pacakge as inspiration) which has functions for performing dynamic prediction using our framework
5. Implement a brief simulation study showing both that the approach works

6. Deriving unbiased scores

As in our first paper (attached), the scores we obtain from the approach you're using are biased. We can show that the scores obtained from the local models (i.e. Step 3 in the fastGFPCA approach) are a linear function of the true scores.
1) Derive the result that the binned scores are linear in the true scores (I have already derived this result, but you should attempt to do this yourself.)
- Attempt to find an analytic solution for the bias factor. This may be challenging, so don't spend too much time here, but it would be an amazing result if we could do so

2) Assuming you cannot find an analytic solution for the bias factor, attempt to estimate this factor yourself. To do so implement the full fastGFPCA approach (i.e. Step 4, where we re-estimate the scores), then:
- Take the estimated scores for each subject from step 4 and and regress them on the scores from step 3 using linear regression. Do this separately for each eigenfunction
-  Identify and report the estimated coefficients and R^2 for each eigenfunction separately
- Use these models to create a de-biasing model for the Step 3 scores
Show these scores are closer (lower MSE) to the true scores than those obtained from Step 3, and that these result in lower MSE for the latent subject specific functoions
- Incorporate this result into the dynamic prediction framework you developed per the instructions above

7. FPCA eigenfunctions: how to extend to the original grid? Interpolation or projection?

Thoughts on 1:

1. If we'd like to make predictions on more than one sample, we may be able to fit GLMM on test samples and predict $\eta(t)$
2. Do a different set of derivation with MLE+EB like [this paper](https://www.jstor.org/stable/pdf/2533715.pdf?refreqid=excelsior%3A135dd765c41e4b46eb39bb6ef7a71053&ab_segments=&origin=&acceptTC=1)

>>>>

---
title: "fGFPCA dynamic prediction (In-sample)"
output: 
  html_document:
    toc: true
    toc_float: 
      collapsed: false
    toc_depth: 4
date: "2022-11-03"
---

@@

library(here)
library(tidyverse)
library(ggplot2)
theme_set(theme_minimal())
library(fcr)
library(refund)
library(lme4)
library(ggpubr)
library(gridExtra)
set.seed(1025)
```

@@

<span style="color: red;">When fitting local GLMMs, it seems that the intervals are essentially independent. We do not consider the effect of neighborhood intervals on a specific interval. This means that correlation between repeated measures from the same subject is only introduced by the latent function.</span>

@@

in_pred_clean <- function(subjID=1, mid=mid){
  # make predictions
  pred_t195 <- in_samp_dyn_pred(fpca_fit, df_pred_unique %>% filter(id==subjID & bin<=195))
  pred_t395 <- in_samp_dyn_pred(fpca_fit, df_pred_unique %>% filter(id==subjID & bin<=395))
  pred_t595 <- in_samp_dyn_pred(fpca_fit, df_pred_unique %>% filter(id==subjID & bin<=595))
  # put predicted function and standard error in to a data frame
  # calculate confidence interval
  df_pred_t195 <- data.frame(mid,
                             eta=pred_t195$eta_pred,
                             eta_std = pred_t195$eta_std_var) %>%
    mutate(eta_lower = eta-eta_std, 
             eta_upper = eta+eta_std)

  df_pred_t395 <- data.frame(mid,
                             eta=pred_t395$eta_pred,
                             eta_std = pred_t395$eta_std_var) %>%
    mutate(eta_lower = eta-eta_std, 
             eta_upper = eta+eta_std)

  df_pred_t595 <- data.frame(mid,
                             eta=pred_t595$eta_pred,
                             eta_std = pred_t595$eta_std_var) %>%
    mutate(eta_lower = eta-eta_std, 
             eta_upper = eta+eta_std) 
  # remove observed part
  df_pred_t195[df_pred_t195$mid<=195, c("eta", "eta_std", "eta_lower", "eta_upper")] <- NA
  df_pred_t395[df_pred_t195$mid<=395, c("eta", "eta_std", "eta_lower", "eta_upper")] <- NA
  df_pred_t595[df_pred_t195$mid<=595, c("eta", "eta_std", "eta_lower", "eta_upper")] <- NA
  # put into one data frame
  df_pred <- bind_rows(df_pred_t195, df_pred_t395, df_pred_t595, .id="type") %>%
    mutate(type = factor(type, levels=1:3, 
                         labels = c("Up to 195", "Up to 395", "Up to 595")))

  return(df_pred)
}

@@

```{r}
# Try on subjects 1-4
in_pred_subj1 <- in_pred_clean(1, mid) %>% 
  left_join(df_true_eta %>% filter(id==1), by = "mid") 
in_pred_subj2 <- in_pred_clean(2, mid) %>% 
  left_join(df_true_eta %>% filter(id==2), by = "mid")
in_pred_subj3 <- in_pred_clean(3, mid) %>% 
  left_join(df_true_eta %>% filter(id==3), by = "mid") 
in_pred_subj4 <- in_pred_clean(4, mid) %>% 
  left_join(df_true_eta %>% filter(id==4), by = "mid") 
```

```{r, fig.width=12, fig.height=4}
in_pred_subj1 %>%
  ggplot()+
  geom_line(aes(x=mid, y = true), col = "red")+
  geom_line(aes(x=mid, y = eta), na.rm = T)+
  geom_line(aes(x=mid, y=eta_lower), linetype = "dashed", na.rm=T)+
  geom_line(aes(x=mid, y = eta_upper), linetype = "dashed", na.rm=T)+
  labs(title = "Subject 1", x = "bin", y="")+
  facet_wrap(~type)

in_pred_subj2 %>% 
  ggplot()+
  geom_line(aes(x=mid, y = true), col = "red")+
  geom_line(aes(x=mid, y = eta), na.rm = T)+
  geom_line(aes(x=mid, y=eta_lower), linetype = "dashed", na.rm=T)+
  geom_line(aes(x=mid, y = eta_upper), linetype = "dashed", na.rm=T)+
  labs(title = "Subject 2", x = "bin", y="")+
  facet_wrap(~type)

in_pred_subj3 %>% 
  ggplot()+
  geom_line(aes(x=mid, y = true), col = "red")+
  geom_line(aes(x=mid, y = eta), na.rm = T)+
  geom_line(aes(x=mid, y=eta_lower), linetype = "dashed", na.rm=T)+
  geom_line(aes(x=mid, y = eta_upper), linetype = "dashed", na.rm=T)+
  labs(title = "Subject 3", x = "bin", y="")+
  facet_wrap(~type)

in_pred_subj4 %>% 
  ggplot()+
  geom_line(aes(x=mid, y = true), col = "red")+
  geom_line(aes(x=mid, y = eta), na.rm = T)+
  geom_line(aes(x=mid, y=eta_lower), linetype = "dashed", na.rm=T)+
  geom_line(aes(x=mid, y = eta_upper), linetype = "dashed", na.rm=T)+
  labs(title = "Subject 4", x = "bin", y="")+
  facet_wrap(~type)
```

@@

### In-/Out-of-sample prediction of latent function

@@

```{r}
# clean out-of-sample prediction results
out_pred_clean <- function(subjID, mid=mid){
  # Prediction
  pred_t195 <- out_samp_dyn_pred(df_new = df %>% filter(id==subjID & sind_inx<=195) %>%
                                  select(-eta_i),
                                fpca_fit = fpca_fit)
  pred_t395 <- out_samp_dyn_pred(df_new = df %>% filter(id==subjID & sind_inx<=395) %>%
                                  select(-eta_i),
                                fpca_fit = fpca_fit)
  pred_t595 <- out_samp_dyn_pred(df_new = df %>% filter(id==subjID & sind_inx<=595) %>%
                                  select(-eta_i),
                                fpca_fit = fpca_fit)

  # clean
  df_pred <- data.frame(mid, 
                        pred_t195=pred_t195$eta_pred,
                        pred_t395=pred_t395$eta_pred, 
                        pred_t595=pred_t595$eta_pred)
  df_pred[df_pred$mid<=195, "pred_t195"] <- NA
  df_pred[df_pred$mid<=395, "pred_t395"] <- NA
  df_pred[df_pred$mid<=595, "pred_t595"] <- NA

  df_pred <- df_pred %>% pivot_longer(starts_with("pred_t"), 
                                      names_to = "type", values_to = "eta") %>%
    mutate(type = factor(type, levels=c("pred_t195", "pred_t395", "pred_t595"),
                         labels=c("Up to 195", "Up to 395", "Up to 595")))

  return(df_pred)
}

@@

```{r}
out_pred_subj1 <- out_pred_clean(1, mid) %>% 
  left_join(df_true_eta %>% filter(id==1), by = "mid")
out_pred_subj2 <- out_pred_clean(2, mid) %>% 
  left_join(df_true_eta %>% filter(id==2), by = "mid")
out_pred_subj3 <- out_pred_clean(3, mid) %>% 
  left_join(df_true_eta %>% filter(id==3), by = "mid")
out_pred_subj4 <- out_pred_clean(4, mid) %>% 
  left_join(df_true_eta %>% filter(id==4), by = "mid")
```

```{r, fig.width=8, fig.height=4}
# compare point prediction with in-sample
# subject 1
grid.arrange(
  out_pred_subj1 %>% 
    group_by("type") %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col = "black")+
    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 1 out-of-sample"),

  in_pred_subj1 %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col="black")+
    geom_line(aes(x=mid, y=eta, col=type, group=type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 1 in-sample"),
  nrow=1)

# subject 2
grid.arrange(
  out_pred_subj2 %>% 
    group_by("type") %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col = "black")+
    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 2 out-of-sample"),

  in_pred_subj2 %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col="black")+
    geom_line(aes(x=mid, y=eta, col=type, group=type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 2 in-sample"),
  nrow=1)

# subject 3
grid.arrange(
  out_pred_subj3 %>% 
    group_by("type") %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col = "black")+
    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 3 out-of-sample"),

  in_pred_subj3 %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col="black")+
    geom_line(aes(x=mid, y=eta, col=type, group=type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 3 in-sample"),
  nrow=1)

# subject 4
grid.arrange(
  out_pred_subj4 %>% 
    group_by("type") %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col = "black")+
    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 4 out-of-sample"),

  in_pred_subj4 %>%
    ggplot()+
    geom_line(aes(x=mid, y=true), col="black")+
    geom_line(aes(x=mid, y=eta, col=type, group=type), linetype="dashed", na.rm = T,
              show.legend = FALSE)+
    labs(x="Binned time", title = "Subject 4 in-sample"),
  nrow=1)

@@

### In-/Out-of-sample PC scores

<span style="color: red;">Doesn't look biased? </span>
<span style="color: red;">Perhaps because the function value range is so small? </span>

```{r}
# true scores:xi

# scores
in_score <- matrix(NA, nrow=N, ncol=4)
out_score <- matrix(NA, nrow=N, ncol=4)
for(i in 1:N){
  in_score[i, ] <- in_samp_dyn_pred(fpca_fit=fpca_fit, 
                   df_part=df_pred_unique %>% filter(id==i & bin<=595))$score
  out_score[i,] <- out_samp_dyn_pred(df %>% filter(id==i & sind_inx<=595) %>%
                                  select(-eta_i),
                                  fpca_fit=fpca_fit)$score_out
}

# bias
in_bias <- in_score-xi
out_bias <- out_score-xi
```

```{r, fig.height=4, fig.width=8}
grid.arrange(
  out_bias %>% data.frame() %>%
    pivot_longer(1:4, names_to = "PC", values_to = "Bias") %>%
    ggplot()+
    geom_boxplot(aes(x=PC, y=Bias))+
    stat_summary(aes(x=PC, y=Bias), fun = mean, geom = "point", col = "red")+
    labs(title = "Out-of-sample"),

  in_bias %>% data.frame() %>%
    pivot_longer(1:4, names_to = "PC", values_to = "Bias") %>%
    ggplot()+
    geom_boxplot(aes(x=PC, y=Bias))+
    stat_summary(aes(x=PC, y=Bias), fun = mean, geom = "point", col = "red")+
    labs(title = "In-sample"), 
nrow=1)

@@

@@

- Interval prediction for MLE
- FPCA eigenfunctions: how to extend to the original grid? Interpolation or projection?
